---
title: "Lactate Correlation Paper: Figure 3"
author: "John Santiago"
date: "2022-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE, message = FALSE}

library(edgeR)
library(gplots)
library(heatmaply)

countdata=read.csv("/Users/johncsantiago/Documents/GitHub/SandersLab/MGH_2020/R Scripts/EdgeR/MGH_AllRawCount.csv",row.names=1)
groups=read.csv("/Users/johncsantiago/Documents/GitHub/SandersLab/MGH_2020/R Scripts/EdgeR/MGH_metadata.csv",row.names=1)
annot =read.csv("https://raw.githubusercontent.com/johncsantiago/SData/master/Apoptosis%20Code/AnnotationData.csv",row.names=1)

##normalize data
countdata=countdata[,row.names(groups)]
x <- countdata
group <- factor(groups$Group)
y <- DGEList(counts=x,group=group)
keep <- filterByExpr(y)
y <- y[keep,,keep.lib.sizes=FALSE]
z <- calcNormFactors(y, method = "TMM")
##normalized cpm
cpmdata=cpm(z)

##Liver Trait Data 
LTD= read.csv("/Users/johncsantiago/Documents/GitHub/Lactate-Correlation-Manuscript/Data/LiverTraitData.csv")
##FunctionalGroups
FG=c(rep("Adequate Function",5),rep("Intermediate Function",2),rep("Low Function",3))
names(FG)=c("FV12","LV12","LV13","FV13","LV11","FN11","FN12","LN11","FN13","LN13")

LacData=LTD$alactate
names(LacData)=LTD$libraryID
LacData=LacData[LTD$time==3]
LacData=LacData[paste0(substring(names(FG),1,2),substring(names(FG),4,4))]
names(LacData)=paste0(substring(names(LacData),1,2),"1",substring(names(LacData),4,4))
cpm.0H=cpmdata[,names(FG)]
cpm.0H=cpm.0H[apply(cpm.0H,1,sum)>0,]

trait.cor=function(gene.data){
  calc.cor=cor.test(LacData,gene.data)
  return(calc.cor[[4]])
}

trait.cor.p=function(gene.data){
  calc.cor=cor.test(LacData,gene.data)
  return(calc.cor[[3]])
}

cor.data=apply(cpm.0H,1,trait.cor)
cor.p=apply(cpm.0H,1,trait.cor.p)

```

<br  /><br  /><br  />
***Figure 3A***<br  />
Heatmap of genes that meet positive correlation with lactate criteria<br  />
```{r echo=F, fig.height=9, fig.width=7}

cpm.cutoff=20
cor.cutoff=0.8
sig.cutoff=.05
pos.or.neg="pos"

##trim for genes that meet cpm.cutoff
cpm.genes=apply(cpm.0H,1,max)
cpm.genes=names(cpm.genes[cpm.genes>cpm.cutoff])

##trim for genes that meet sig.cutoff
sig.genes=names(cor.p)[cor.p<sig.cutoff]
  
##trim for genes that meet the cor.cutoff in the right direction
if(pos.or.neg=="pos"){
  cor.genes=names(cor.data)[cor.data>cor.cutoff]
}
if(pos.or.neg=="neg"){
  cor.genes=names(cor.data)[cor.data<cor.cutoff]
}
  
cordata=intersect(cpm.genes,intersect(cor.genes,sig.genes))

hmdata = cpm.0H[cordata,]
hc = hclust(dist(hmdata), "complete")
hmdata=hmdata[hc$order,]  
pos.hmdata=t(scale(t(hmdata)))

heatmaply(pos.hmdata,trace="none",col=RdYlBu(100)[100:1], scale="none",
          dendrogram = "none",Rowv=F,Colv=F,
          cexRow = .75, na.color="grey",
          labRow = annot[row.names(hmdata),"symbol"],
          labCol = c("AF1","AF2","AF3","AF4","AF5","IF1","IF2","LF1","LF2","LF3"),
          main="Positively Correlating Genes")

```


<br  /><br  /><br  />
***Figure 3B***<br  />
Scaled expression of positively correlating genes and lactate levels<br  />
```{r echo=F, fig.height=7, fig.width=7}

##Positively correlating genes: relative expression, mean rel. expr. and lactate
geneset=row.names(pos.hmdata)
mean.scaled=apply(pos.hmdata,2,mean)
var.scaled=apply(pos.hmdata,2,var)

mean.line=function(mean.scaled,var.scaled){
  lines(y=mean.scaled,x=1:length(mean.scaled),type="l",col="black",lwd=2,lty=1)
  i=1
  while(i<=length(var.scaled)){
    y1=mean.scaled[i]+var.scaled[i]
    y2=mean.scaled[i]-var.scaled[i]
    cap1=i-.1
    cap2=i+.1
    lines(x=c(i,i),y=c(y1,y2),col="black")
    lines(x=c(cap1,cap2),y=c(y1,y1),col="black")
    lines(x=c(cap1,cap2),y=c(y2,y2),col="black")
    i=i+1
  }
}

plot(y=c(rep(1,length(LacData))),x=1:length(LacData),type="l",main="Lactate levels and positive correlating genes (relative expression)",xlab="Sample",ylab="Relative Gene Expression",xaxt="n",col=NA,lwd=2,lty=1,ylim=c(-1,1.75))
axis(side=1, at=1:length(LacData), labels=names(LacData), cex.axis=1,las=2)

i=1
while(i<=length(geneset)){
  lines(pos.hmdata[i,],col=alpha("lightgrey", 0.6))
  i=i+1
}

mean.line(mean.scaled,var.scaled)
mean.cor=cor(LacData,mean.scaled)
mean.p=cor.test(LacData,mean.scaled)[[3]]


  
par(new=T)
plot(y=LacData,x=1:length(LacData),type="l",yaxt="n",xaxt="n",col="red3",lty=2,lwd=2,xlab="",ylab="",ylim=c(0,22.5))
axis(side=4)

mtext("Lactate mmol/L", side = 4, line = 3)


legend("topleft",legend=c("Pos. Cor. Genes (mean relative expression)","3H Lactate",paste("r = ",signif(mean.cor,3),sep=""),paste("p = ",signif(mean.p,3),sep="")),col=c("firebrick","black",NA,NA),lty=c(1,1,2),bty = "n",xpd=NA,cex=.7)



```


<br  /><br  /><br  />
***Figure 3C***<br  />
Linear model of mean scaled expression of positively correlating genes and lactate levels<br  />
```{r echo=F, fig.height=7, fig.width=7}
##Pos. Cor. Genes
mean.scaled=apply(pos.hmdata,2,mean)

linearmodel=lm(LacData~mean.scaled)
R2=cor(LacData,mean.scaled)^2
plot(y=LacData,x=mean.scaled,type="p",main="Lactate~Mean Scaled Data",xlab="Mean Scaled Data",ylab="Lactate mmol/L",col="black",lwd=2,lty=1)
abline(a=linearmodel[[1]][1],b=linearmodel[[1]][2],lty=2)
if(linearmodel[[1]][2]>0){
  text(x=(min(mean.scaled)+(max(mean.scaled)-min(mean.scaled))*.25),y=max(LacData)*.9,
       labels=paste("lm=",signif(linearmodel[[1]][2],3),"x + ",signif(linearmodel[[1]][1],3), "\nR^2 = ",signif(R2,3),sep=""))
}

if(linearmodel[[1]][2]<0){
  text(x=(min(mean.scaled)+(max(mean.scaled)-min(mean.scaled))*.75),y=max(LacData)*.9,
       labels=paste("lm=",signif(linearmodel[[1]][2],3),"x + ",signif(linearmodel[[1]][1],3), "\nR^2 = ",signif(R2,3),sep=""))
}
```

<br  /><br  /><br  />
***Figure 3D***<br  />
Heatmap of genes that meet positive correlation with lactate criteria<br  />
```{r echo=F, fig.height=9, fig.width=7}

cpm.cutoff=20
cor.cutoff=0.8
sig.cutoff=.05
pos.or.neg="neg"

##trim for genes that meet cpm.cutoff
cpm.genes=apply(cpm.0H,1,max)
cpm.genes=names(cpm.genes[cpm.genes>cpm.cutoff])

##trim for genes that meet sig.cutoff
sig.genes=names(cor.p)[cor.p<sig.cutoff]
  
##trim for genes that meet the cor.cutoff in the right direction
if(pos.or.neg=="pos"){
  cor.genes=names(cor.data)[cor.data>cor.cutoff]
}
if(pos.or.neg=="neg"){
  cor.genes=names(cor.data)[-cor.data>cor.cutoff]
}
  
cordata=intersect(cpm.genes,intersect(cor.genes,sig.genes))

hmdata = cpm.0H[cordata,]
hc = hclust(dist(hmdata), "complete")
hmdata=hmdata[hc$order,]  
neg.hmdata=t(scale(t(hmdata)))
heatmaply(neg.hmdata,trace="none",col=RdYlBu(100)[100:1], scale="none",
          dendrogram = "none",Rowv=F,Colv=F,
          cexRow = .5, na.color="grey",
          labRow = annot[row.names(hmdata),"symbol"],
          labCol = c("AF1","AF2","AF3","AF4","AF5","IF1","IF2","LF1","LF2","LF3"),
          main="Negatively Correlating Genes")

```


<br  /><br  /><br  />
***Figure 3E***<br  />
Scaled expression of negatively correlating genes and lactate levels<br  />
```{r echo=F, fig.height=7, fig.width=7}

##Negatively correlating genes: relative expression, mean rel. expr. and lactate
geneset=row.names(neg.hmdata)
mean.scaled=apply(neg.hmdata,2,mean)
var.scaled=apply(neg.hmdata,2,var)

mean.line=function(mean.scaled,var.scaled){
  lines(y=mean.scaled,x=1:length(mean.scaled),type="l",col="black",lwd=2,lty=1)
  i=1
  while(i<=length(var.scaled)){
    y1=mean.scaled[i]+var.scaled[i]
    y2=mean.scaled[i]-var.scaled[i]
    cap1=i-.1
    cap2=i+.1
    lines(x=c(i,i),y=c(y1,y2),col="black")
    lines(x=c(cap1,cap2),y=c(y1,y1),col="black")
    lines(x=c(cap1,cap2),y=c(y2,y2),col="black")
    i=i+1
  }
}

plot(y=c(rep(1,length(LacData))),x=1:length(LacData),type="l",main="Lactate levels and negative correlating genes (relative expression)",xlab="Sample",ylab="Relative Gene Expression",xaxt="n",col=NA,lwd=2,lty=1,ylim=c(-1.5,1.75))
axis(side=1, at=1:length(LacData), labels=names(LacData), cex.axis=1,las=2)

i=1
while(i<=length(geneset)){
  lines(neg.hmdata[i,],col=alpha("lightgrey", 0.6))
  i=i+1
}

mean.line(mean.scaled,var.scaled)
mean.cor=cor(LacData,mean.scaled)
mean.p=cor.test(LacData,mean.scaled)[[3]]


  
par(new=T)
plot(y=LacData,x=1:length(LacData),type="l",yaxt="n",xaxt="n",col="red3",lty=2,lwd=2,xlab="",ylab="",ylim=c(0,22.5))
axis(side=4)

mtext("Lactate mmol/L", side = 4, line = 3)


legend("topright",legend=c("Neg. Cor. Genes (mean relative expression)","3H Lactate",paste("r = ",signif(mean.cor,3),sep=""),paste("p = ",signif(mean.p,3),sep="")),col=c("firebrick","black",NA,NA),lty=c(1,1,2),bty = "n",xpd=NA,cex=.7)

```

<br  /><br  /><br  />
***Figure 3F***<br  />
Linear model of mean scaled expression of negatively correlating genes and lactate levels<br  />
```{r echo=F, fig.height=7, fig.width=7}
##Neg. Cor. Genes
mean.scaled=apply(neg.hmdata,2,mean)

linearmodel=lm(LacData~mean.scaled)
R2=cor(LacData,mean.scaled)^2

plot(y=LacData,x=mean.scaled,type="p",main="Lactate~Mean Scaled Data",xlab="Mean Scaled Data",ylab="Lactate mmol/L",col="black",lwd=2,lty=1)
abline(a=linearmodel[[1]][1],b=linearmodel[[1]][2],lty=2)
if(linearmodel[[1]][2]>0){
  text(x=(min(mean.scaled)+(max(mean.scaled)-min(mean.scaled))*.25),y=max(LacData)*.9,
       labels=paste("lm=",signif(linearmodel[[1]][2],3),"x + ",signif(linearmodel[[1]][1],3), "\nR^2 = ",signif(R2,3),sep=""))
}

if(linearmodel[[1]][2]<0){
  text(x=(min(mean.scaled)+(max(mean.scaled)-min(mean.scaled))*.75),y=max(LacData)*.9,
       labels=paste("lm=",signif(linearmodel[[1]][2],3),"x + ",signif(linearmodel[[1]][1],3), "\nR^2 = ",signif(R2,3),sep=""))
}
```